\documentclass[a4paper]{article}

%% Language and font encodings
\usepackage[english]{babel}
\usepackage[utf8x]{inputenc}
\usepackage[T1]{fontenc}

%% Sets page size and margins
\usepackage[a4paper,top=3cm,bottom=2cm,left=3cm,right=3cm,marginparwidth=1.75cm]{geometry}

%% Useful packages
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage[colorinlistoftodos]{todonotes}
\usepackage[colorlinks=true, allcolors=blue]{hyperref}

\title{Odi Cricket Prediciton}
\author{Aditya Yadav\\ 
\texttt{adiyadav.1729@gmail.com }\and
Aniket Dhawad\\
\texttt{aniketdhawadiitg1@gmail.com} \and
Vikash Goyal \\
\texttt{vikashgoyal2701@gmail.com}}



\begin{document}
\maketitle

\begin{abstract}
The aim of this study was to investigate to what degree a machine learning model can capture outcome of match depending on team's composition and strength. The target competition was the World Cup 2019.   Basic features comprising of batting and bowling metrics was used along with features capturing the impact and quality of players. Team's batting and bowling strength was created based on playing squad for a match and venue where the match was played. 
The model was tested with different training dataset based on timeline. Result showed how greatly odi crickets have changed with time. 
\end{abstract}

\section{Introduction}

One Day International(ODI) Cricket is one of the three formats which are played at International Level. International Cricket Council(ICC), governing body of cricket, has  currently 12  Full Team Members and 92 Associate Members. 
ODI cricket started in 1971 as second form of cricket after test cricket to cater entertainment to larger set of spectators.In this format, both teams are allowed to play for fixed overs and the one with more runs is declared winner. Originally started with 60 overs and white jerseys, ODI cricket has gone through various transition over years  
to stay relevant and consistently entertaning. 50 overs , powerplay overs, field restrictions, batting friendly grounds are among few significant changes that has played role in its evolution.

Cricket World Cup, the most important tournament in this format generally played every 4 years, with preliminary qualification rounds leading up to a finals tournament. The tournament is one of the world's most viewed sporting events and is considered the "flagship event of the international cricket calendar" by the ICC.  World cup 2019 has 10 teams participating. In league round all 10 teams would be playing against each other. Top 4 would march into semi- finals. Winner of the semi-finalist would be playing for the coveted title. There would be 45 league matches, 2 semi-finals and 1 finals. This paper aimed at predicting the result of 48 matches.

We collected data from cricinfo and relianceiccranking. Over 100 features were created which included players and teams features as well.  Pearson correlation, chi square test and recursive feature elimination were used for feature elimination. The selected features were used as inputs to three different classification algorithms: logistic regression,
random forest, boosted decision trees. Ensemble of weak classifiers were also tested. Best performing model was selected as final one to predict the outcome.


\section{Methodolgy}

\subsection{Data}

Data was scrapped from www.espncricinfo.com. Firstly, list of all matches played was scrapped. Secondly, list of all players played for different country was scrapped. Note that data had match\_id, player\_id which was key in getting different information
at match and player level respectively. We used www.relianceiccrankings.com to collect information of player batting and bowling ranking at the start of every month from 1971 till date. 

Therefore, for every match we had batting and bowling scorecard, player details at match level and their ranking. 



\subsection{Feature Engineering}

\subsubsection{
Basic Metrics
}
Players Batting Average, Strike rate, Bowling Average , Bowling Economy, Bowling Strike Rate was considered just before a match for which features were generated. \newline

To compare like to like players from two teams, players role was identified.  That is, features were arranged such that openers from a team would be compared against openers from its opposition. Similarly, for bowleres , first change bowlers would be observed against similar role bowler from opposition team

\subsubsection{
Impact Factor
}

This factor was created to compare players based on its quality rather than amount of runs scored/wickets taken.
Impact features take into account the quality of opposition bowlers against which runs scored. A batsman scoring runs against a bowler who has a bowling average of 25 will have higher impact score compare to scoring runs against a bowling average of 35.


Similarly, Bowlers taking higher order wickets are given higher weightage for batsmen dismissed.  This could lead to a situation where bowler with a high average(or poor strike rate) can get higher weightage on wickets. Hence,  strike rate was taken into factor.
\newline
\[Impact Factor,Bowler = \frac{weighted sum of wickets}{wickets * strike rate}
 

\newline
\begin{table}
\centering
\begin{tabular}{c|c}
Wicket Order & Wicket Weight\\\hline
 1 & 1.5\\
2  & 1.4 \\
3 & 1.5 \\
4 & 1.3 \\
5 & 1.2\\
6 & 0.9 \\
7 & 0.7 \\
8 & 0.5 \\
9 & 0.5\\
10  & 0.3 \\
11 & 0.2 \\
\end{tabular}
\caption{\label{tab:widgets}Wickets Order and Weightage .}
\end{table}

Matt Henry till Feb 2019 had close to 45\% top order wickets(1-3) out of 78 batsmen he dismissed. Lasith Malinga on other hand had close to 38\% of top order wickets  in his 322 wickets. Henry's took wickets at strike rate of 29 whereas Malinga's took struck on every 32 balls. Each of these features was taken to calculate their impact factor.  

\begin{table}
\centering
\begin{tabular}{c|c|c|c|c|c}
Players & Matches & Wkts & Weighted Wkts & SR & Impact\\\hline
 Matt Henry & 43 & 78 & 91.2 & 28.7 & 1686\\
 Lasith Malinga & 218 & 322 & 329 & 32.6 & 990
\end{tabular}
\caption{\label{tab:widgets}Impact Factor : Bowler.}
\end{table}

Top 3 batsman with highest impact before  2019 world cup\newline 
Top 3 batsman with highest impact before 2015 world cup


\subsubsection{
Form Factor
}

{Batting}: \newline
While comparing batsmen , generally batting average is used a metric to identify the batsmen in form. Though, it is reasonable to comparable batsmen on average when a considerable time period is taken into account. However, it does not give clear picture of batsmen form in last few( say 10) matches. For instance, a batsmen with score sequence  120,0,30,40,10  another with score 60,45,45,35 have batting average of 50. Though it is fair to assume second batsmen is more consistent. 

In addition to runs scored, rate at which they are scored also play a crucial role in fate of a match. This is also taken in consideration to determine batsmen's recent form.

For instance, before England's league match against India in world cup 2019,  Ben Stokes and Bairstow had following streaks. Bairstow is more of a consistent batsmen(2nd row) with a standard deviation of 33 , however comparing Stokes recent number with overall number, he is actually punching above his weight and thus had a higher form factor.

\begin{table}
\centering
\begin{tabular}{c|c|c|c|c|c}
Runs & Balls & Standard Deviation & Rec Avg * SR & Career Avg * SR  & Form Factor\\\hline
 89,82*,2,10*,6 & 115,89,6,6,7 & 43 & 5340 & 3672 & 2.1\\
 27,0,90,45,91  & 39,1,99,46,50 & 33 & 3861 & 4842 & 1.02
\end{tabular}
\caption{\label{tab:widgets}Form Factor.}
\end{table}


{Bowling}: \newline
This metric was calculated to capture the form and consistency of the bowler in the recent few matches before the actual match. Bowler's economy, strike rate and average in the last 10 matches was taken into consideration while calculating this metric. 
Example:
Show data of last 10 matches of a particular inform bowler and a out of form bowler.

\subsubsection{Team Strength
}

For every team, batting and bowling strength was calculated by considering the ratings of players of the team as per the ICC world rankings. The batting strength was calculated by taking the average rating of top 6 batsmen and bowling strength by taking the average rating of top 5 bowlers. We assumed the minimum rating  for player who was not present in the top 100 list.

Pitch Weighted Team Strength:
In order to take into the impact of the pitch, we tried to quantify whether the pitch was a high scoring or a low scoring one. We did this by calculating the average runs that are scored on a particular pitch. If the pitch was high scoring one, we increased the batting strength of a team and reduce the bowling strength of a team by a particular factor..

Rationale behind this feature was that high scoring pitches are reasonably flat and give advantage to batsmen and reduce the advantage to bowlers.


\subsection{Modelling}
 
 
\subsubsection{Feature Selection
}

 Over 120 features were created.  Pearson correlation scores were calculated to identify and hence remove correlation among features. Features such as Batting average and strike rate was highly correlated and they were multiplied to create another feature. Similarly, bowling average, strike rate and economy was clubbed to make one feature. Also, features like team rate was introducing bias and hence this was also not shortlisted in final list of feature.
 
\subsubsection{Train and Test Data
}
One day cricket had undergone immense change since its inception. Hence selection of training data became an important factor. On experimenting with training data of different period , we identified post 2001 period as best suited training data for our model. Post 2017 one day matches were taken as test data.

\subsubsection{Modelling
}

Finally 78 features were selected were model. We tried logistic regression , Extreme Gradient Boosting and Random Forest to generate predictions. 
Since, sports data is more likely to have noise, Random Forest is suitable as it is robust in dealing with noise. Extreme Gradient Boosting , owing to its characteristics of capturing even noise, was over-fitting and hence resulted in poor performance. 
 
 Random Forest gave an accuracy of 70\% in test data.

\begin{thebibliography}{9}
\bibitem{texbook}
www.cricinfo.com
\bibitem{site}
http://www.relianceiccrankings.com/
\bibitem{paper}
Using Machine Learning to Predict the Outcome of English
County twenty over Cricket Matches - Stylianos Kampakis, 
William Thomas





\end{thebibliography}

\end{document}

